---
title: "Project Draft: Figure 1"
author: "Caitlin Bolz, Matt Voss, Steven Xia"
output: pdf_document
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(readr)
library(dplyr)
```


```{r include=FALSE}
UberandWeather = read_csv("UberandWeather.csv")
```

```{r include=FALSE}
sum<-UberandWeather %>%
  filter(month == 5) %>% 
  group_by(day) %>% 
  summarize(pickup=n(), temp = mean(temperature))

sum$day = as.numeric(sum$day)
```


```{r echo=FALSE}
ggplot(sum, aes(x = day, y = pickup, fill = temp)) +
  geom_col() +
  scale_fill_gradient(low = "white", high = "red") +
  labs(
    title = "Number of Uber Pickups in NYC in May 2014",
    x = "Day of the Month",
    y = "Number of Pickups",
    fill = "Temperature") +
  scale_x_continuous(breaks = waiver(), n.breaks = 6)
```

Our project analyzes the specific factors that influence ‘Uber’ app usage within New York City, specifically looking at May 2014. Figure1 implements a bar graph visualization that helps us further explore our overarching question. The total number of Uber pickups within a given day is on the y-axis, and each day of the month across the x-axis. The color of each bar is encoded as the temperature in Fahrenheit of each day, with lighter colors representing colder temperatures. 


As shown in this figure, the number of Uber pickups seem to oscillate in a consistent pattern. The lowest usage point for each week happens around Sunday/Monday. Uber usage rate then rises steadily each day until usage peaks on Thursdays/Fridays. This pattern is apparent across all four weeks of May 2014. It is shown that usage peaks every week, at the end of each work week (Thursday/Friday). This makes sense because the weekend is when people usually have time for leisure activities, that might be out of their daily routine and require alternate transportation.
